---
title: "PhyloGLM of secondary woodiness"
author: "Rutger Vos (@rvosa)"
date: "21-5-2020"
output:
  html_document: default
  pdf_document: default
---

## Loading data

A data frame of OMI-transformed GIS values and a classification into secondary
woodiness (or not) is available in the '../data' folder, next to a pre-processed
version of the ALLMB tree of Smith & Brown. Here we load these into the 
environment:

```{r data}
library(ape)
load( file = '../data/woodyData.rda' )
load( file = '../data/woodyTree.rda' )
plot(tree_woody, show.tip.label = F)
```

## Variable selection

The data frame has 42 columns, one of which is a binary state (secondary woody),
the others are continuous-valued, OMI-transformed GIS values. Out of these, a
selection has been made of variables that are not (very) collinear (using VIF)
but include variables that are hypothesized to predict secondary woodiness. The 
next step is to figure out which of these remaining variables actually 
contribute to a generalized linear model to predict secondary woodiness. Here we
do this using a function defined in the '../R' folder:

```{r phyloglmstep}
library(phylolm)
source('../R/phyloglmstep.R')

# the remaining variables after guided VIF collinearity correction
woody_formula_24a <- woody ~ Aspect + BulkDensity + ClayPercentage + 
    monthCountByTemp10 + PETDriestQuarter + PETWettestQuarter + OrganicCarbon + 
    PhCaCL + Slope + bio2 + bio5 + bio14 + bio15 + bio18 + bio19

# And now do the call:
phyloglmstep(woody_formula_24a, starting.formula = NULL, data=GIS_woody, 
    phy=tree_woody, method="logistic_MPLE", direction = "forward", trace = 0,
    btol = 100, log.alpha.bound = 8, start.beta=NULL, start.alpha=NULL, 
    boot = 0, full.matrix = TRUE, k = 2 )
```

## Run the phyloglm

With these selected variables we can now construct a model and run the phyloglm:

```{r phyloglm}
# construct the model formula
myModel <- woody ~ 1 + monthCountByTemp10 + PETWettestQuarter + OrganicCarbon +
    PhCaCL + Slope + bio5 + bio14

# run the phyloglm
myResult <- phyloglm(myModel, data = GIS_woody, phy = tree_woody, 
    method = "logistic_MPLE", btol = 100, log.alpha.bound = 8)

summary(myResult)
```

## Plot results

To interpret the results we can make boxplots of the values for woody and non-
woody:

```{r boxplots}
library(dplyr)

GIS_filtered_non_woody <- GIS_woody %>%
    select( woody, monthCountByTemp10, PETWettestQuarter, OrganicCarbon,
    PhCaCL, Slope, bio5, bio14) %>%
    subset(woody == 0)

GIS_filtered_woody <- GIS_woody %>%
    select( woody, monthCountByTemp10, PETWettestQuarter, OrganicCarbon,
    PhCaCL, Slope, bio5, bio14) %>%
    subset(woody == 1)

boxplot(GIS_filtered_woody$monthCountByTemp10,GIS_filtered_non_woody$monthCountByTemp10,
    main = "number of months with mean temp greater than 10 degrees C",
    at = c(1,2),
    names = c("secondary woody", "herbaceous"),
    las = 2,
    col = c("orange","red"),
    border = "brown",
    horizontal = TRUE,
    notch = TRUE
)

boxplot(GIS_filtered_woody$PETWettestQuarter,GIS_filtered_non_woody$PETWettestQuarter,
    main = "mean monthly potential evapotranspiration of wettest quarter",
    at = c(1,2),
    names = c("secondary woody", "herbaceous"),
    las = 2,
    col = c("orange","red"),
    border = "brown",
    horizontal = TRUE,
    notch = TRUE
)

boxplot(GIS_filtered_woody$OrganicCarbon,GIS_filtered_non_woody$OrganicCarbon,
    main = "soil organic carbon content",
    at = c(1,2),
    names = c("secondary woody", "herbaceous"),
    las = 2,
    col = c("orange","red"),
    border = "brown",
    horizontal = TRUE,
    notch = TRUE
)

boxplot(GIS_filtered_woody$PhCaCL,GIS_filtered_non_woody$PhCaCL,
    main = "soil pH",
    at = c(1,2),
    names = c("secondary woody", "herbaceous"),
    las = 2,
    col = c("orange","red"),
    border = "brown",
    horizontal = TRUE,
    notch = TRUE
)

boxplot(GIS_filtered_woody$Slope,GIS_filtered_non_woody$Slope,
    main = "slope",
    at = c(1,2),
    names = c("secondary woody", "herbaceous"),
    las = 2,
    col = c("orange","red"),
    border = "brown",
    horizontal = TRUE,
    notch = TRUE
)

boxplot(GIS_filtered_woody$bio5,GIS_filtered_non_woody$bio5,
    main = "Max temperature of warmest month",
    at = c(1,2),
    names = c("secondary woody", "herbaceous"),
    las = 2,
    col = c("orange","red"),
    border = "brown",
    horizontal = TRUE,
    notch = TRUE
)

boxplot(GIS_filtered_woody$bio14,GIS_filtered_non_woody$bio14,
    main = "Precipitation of driest month",
    at = c(1,2),
    names = c("secondary woody", "herbaceous"),
    las = 2,
    col = c("orange","red"),
    border = "brown",
    horizontal = TRUE,
    notch = TRUE
)
```